Mathematical Aspects of the Capital Market

### 2014-08-13 thesis/cross_correlation_gaussian.tex
Finalized thesis.

\chapter{Covariance Matrix of Gaussian Returns}
\label{chp:Gaussian}
In this chapter we present some results about the covariance
matrix of Gaussian returns. The assumption of Gaussian returns is of
course an over-simplification of reallity, but nevertheless lends some
insight into the problem of elements' and eigenvalues' distributions of
a covariance matrix.
Our primary interest is in the influence that auto-correlations in
returns exert on the covariance matrix. Section \ref{sec:GCC-analytical} and
section \ref{sec:GCC-numerical} discuss how the distributions of the matrix
elements and the eigenvalues are affected by such autocorrelations,
respectively.

\section{Distribution of the Matrix Elements}
\label{sec:GCC-analytical}
In this section we study the distribution of the elements of a
covariance matrix $C = \mtx{RR'}/T$, where
\begin{eqnarray*}
  \mtx R &=&
  \begin{pmatrix}
    r_{11} & r_{12} & \cdots & r_{1T} \\
    r_{21} & r_{22} & \cdots & r_{2T} \\
    \vdots & \vdots & \ddots & \vdots \\
    r_{N1} & r_{N22} & \cdots & r_{NT} \\
  \end{pmatrix}
\end{eqnarray*}
In words, an element $r_{it}$ of $\mtx R$ is the return of asset $i$
at time $t$, with $i=1,\cdots, N$ and $t = 1, \cdots, T$. If each
column of $\mtx R$ follows a zero-mean Gaussian distribution, i.e.
\begin{eqnarray*}
  \begin{pmatrix}
    r_{1t} \\
    \vdots \\
    r_{Nt}
  \end{pmatrix} \sim N(0, \mtx \Sigma)
\end{eqnarray*}
for all $t = 1, \cdots, T$, and none of the return series is
auto-correlated, i.e. $\text{corr}(r_{it}, r_{i,t'}) = 0$ for all $i =
1, \cdots, N$ and $t \neq t'$, then $\mtx{RR'}$ is a Wishart matrix whose
probability density function is well known \cite{Anderson2003}.

>>>>

\chapter{Covariance Matrix of Gaussian Returns}
\label{chp:Gaussian}
In chapter \ref{chp:PriceModels} we have studied \gls{garch} and
\gls{sv} models and seen their power of forecasting future
volatilities. However, we have not considered the important fact that
a financial market comprises many assets and the volatilities of these
assets are correlated to each other in a complicated
manner. Practically useful volatility forecasts require good
understanding of these correlations.

In the literature, covariance matrices of Gaussian and L\'evy
distributed returns have been studied (see e.g. \cite{politi2010,
  Chiani2012, Lalley2013}). However, as is seen in chapter
\ref{chp:PriceModels}, real-world returns are not described by any
particular distribution but rather by stochastic processes that
account for auto-correlations in the returns and the volatilities.

Therefore, the focus of this and the next chapter is on covariance
matrices of realistic return series, and especially covariance matrices
in the case when the return series have considerable
auto-correlations. In particular, we study covariance matrices of
\gls{garch} (1,1) return series in chapter
\ref{chp:CrossCorrelationFat} and show the influence of
auto-correlations on these matrices. But before that, it is useful to
first understand the influence of auto-correlations on covariance
matrices of Gaussian return series. When auto-correlations are absent,
these matrices are called Wishart matrices and have been studied
extensively. The results we obtain in this chapter will provide a
reference to the studies in chapter \ref{chp:CrossCorrelationFat}.

% In this chapter we present some results about the covariance
% matrix of Gaussian returns. The assumption of Gaussian returns is of
% course an over-simplification of reallity, but nevertheless lends some
% insight into the problem of elements' and eigenvalues' distributions of
% a covariance matrix.
% Our primary interest is in the influence that auto-correlations in
% returns exert on the covariance matrix.

In section \ref{sec:GCC-analytical} we discuss how the distributions
of the matrix elements are affected by autocorrelations, and in
section \ref{sec:GCC-numerical} we investigate the distribution of the
eigenvalues.

\section{Distribution of the Matrix Elements}
\label{sec:GCC-analytical}
In this section we study the distribution of the elements of a
covariance matrix $C = \mtx{RR'}/T$, where
\begin{eqnarray*}
  \mtx R &=&
  \begin{pmatrix}
    r_{11} & r_{12} & \cdots & r_{1T} \\
    r_{21} & r_{22} & \cdots & r_{2T} \\
    \vdots & \vdots & \ddots & \vdots \\
    r_{N1} & r_{N22} & \cdots & r_{NT} \\
  \end{pmatrix}
\end{eqnarray*}
and $\mtx R'$ denotes the transpose of $\mtx R$.
In words, an element $r_{it}$ of $\mtx R$ is the return of asset $i$
at time $t$, with $i=1,\cdots, N$ and $t = 1, \cdots, T$. If each
column of $\mtx R$ follows a zero-mean Gaussian distribution, i.e.
\begin{eqnarray*}
  \begin{pmatrix}
    r_{1t} \\
    \vdots \\
    r_{Nt}
  \end{pmatrix} \sim N(0, \mtx \Sigma)
\end{eqnarray*}
for all $t = 1, \cdots, T$, and none of the return series is
auto-correlated, i.e. $\text{corr}(r_{it}, r_{i,t'}) = 0$ for all $i =
1, \cdots, N$ and $t \neq t'$, then $\mtx{RR'}$ is a Wishart matrix whose
probability density function is well known \cite{Anderson2003}.

### 2014-08-13 thesis/price_models.tex
Finalized thesis.

The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a geometric Brownian motion whose increment
$\sigma dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
\end{eqnarray*}
Then by It\^o's lemma, a stochastic differential equation for the
logarithmic price $\ln S$ can be obtained
\begin{eqnarray*}
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$ \cite{Bernt2000}. Therefore,
in a discrete-time model, where the length of each time step $\Delta
t$ is fixed, the return $r_t = \ln S(t) - \ln S(t - \Delta t)$ is
assumed to be Gaussian distributed and have mean and variance that are
functions of $\Delta t$.

@@

``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A \gls{garch}(p, q) model is defined by the following
equation system \cite{Bollerslev86}:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu_t + \epsilon_t \\
  \end{aligned}
\end{equation}
where $r_t$ is the return as mentioned earlier; $p$ and $q$ are
constant integers; $\mu_t$ is the mean process and considered a small
constant for intraday returns. $\epsilon_t$ is called the innovation
of the return and is a variate whose conditional distribution
\footnotemark is Gaussian and has variance $\sigma^2_t$:
\begin{eqnarray*}
  \sigma_t^2 &=& \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
  \sum_{i=1}^p \beta_i \sigma_{t-i}^2
\end{eqnarray*}
\footnotetext{
By conditional distribution we mean the distribution conditional on
the variate's history: The conditional distribution at time $t$ means
the distribution conditional on the history up to time $t-1$
in the discrete-time setting.
}

@@

Figure \ref{fig:nordea_15min_y_qq} shows the normal probability plot of
$y_t$. It is evident from this figure that $y_t$ has fat tails.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=78 255 109
  123]{../pics/nordea2_y_normplot.pdf}
  \caption{\small \it Normal probability plot of residuals of
    log-volatility ($y_t$)}
  \label{fig:nordea_15min_y_qq}
\end{figure}
In addition, a simple calculation reveals that the distribution of
$y_t$ has skewness 0.2988 (shown in table
\ref{tab:nordea_15min_y_moments}).
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691 \\
    \hline
  \end{tabular}
  \caption{\small \it Moments of log-volatility residuals ($y_t$)}
  \label{tab:nordea_15min_y_moments}
\end{table}
Based on this information, we find that $y_t$ can be well
described by a Johnson Su distribution \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  %\vspace{-18mm}
  \centering
    \includegraphics[scale=0.4, clip=true, trim=92 229 116
    133]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{\small \it Log-volatility residuals $y_t$ fitted to a
      Johnson Su distribution. Horizontal: values of $y_t$, denoted x;
      Vertical: $\ln\left(P(y_t < x)\right)$.}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}

@@

First of all, we look at the means and standard deviations of $\ln \sigma^F_t -
\ln \hat{\sigma}_t$, which are listed in table \ref{tab:nordea_2012}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    & \gls{sv} & \gls{garch} & Sample mean \\
    \hline
    $\E(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.0040 & -0.0008 &
    -0.2210 \\
    \hline
    $\text{std}(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.2659 & 0.3011 &
    0.2893 \\
    \hline
  \end{tabular}
  \caption{\small \it Mean and standard Deviation of the forecasts'
    distribution}
  \label{tab:nordea_2012}
\end{table}
It is seen from table \ref{tab:nordea_2012} that, on average, the \gls{sv}
model over-estimates while \gls{garch} under-estimates. In terms of the
standard deviation of $\ln \sigma^F_t - \ln \hat{\sigma}_t$, the \gls{sv}
model wins with a small margin. In contrast, the sample mean forecast
clearly under-estimates the log-volatilities to a large extent --- the
efforts of building models has not been wasted.

@@

Another measure of the forecasts' quality can be the percentage of good
forecasts, where the criterion of ``good'' is defined, respectively,
as the forecast lying within 1\%, 5\%, or 10\% of the corresponding
realized volatility. Table \ref{tab:nordea_2012_good} shows the
respective percentage of the 3 kinds of forecasts. Again in this table
it is seen that the \gls{sv} model gives more accurate forecasts than does
\gls{garch}. For example, defining a ``good'' forecast as one that lies
within 5\% of the realized volatility, the probability of obtaining
such a good forecast is 75\% using the \gls{sv} model, 71\% using the
\gls{garch} model, and only 53\% using the sample mean.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln \hat{\sigma}_t|}$
    & \gls{sv} & \gls{garch} & sample mean \\
    \hline
    1\% & 22\% & 14\% & 11\% \\
    \hline
    5\% & 75\% & 71\% & 53\% \\
    \hline
    10\% & 97\% & 94\% & 89\% \\
    \hline
  \end{tabular}
  % \begin{tabular}{|c|c|c|c|}
  %   \hline
  %   ${\ln \sigma^F_t - \ln \hat{\sigma}_t \over |\ln \hat{\sigma}_t|}$
  %   & \gls{sv} & GARCH & sample mean \\
  %   \hline
  %   1\% & 21.73\% & 13.83\% & 11.36\% \\
  %   \hline
  %   5\% & 74.57\% & 70.62\% & 53.33\% \\
  %   \hline
  %   10\% & 97.28\% & 93.83\% & 89.14\% \\
  %   \hline
  % \end{tabular}
  \caption{\small \it The percentage of ``good'' forecasts when the
    criterion of being good is deviating no more than 1\%, 5\% or
    10\%.}
  \label{tab:nordea_2012_good}
\end{table}

@@

\subsection{The Simplified model}\label{sec:SLV_Symmetric}
In the following we derive the unconditional \gls{pdf} of $r'$, the
de-meaned and rescaled return. Denote this \gls{pdf} $f_{r'}(x)$. Then
the \gls{pdf} of $r$ is $e^{-\bar{v}}f_{r'}(e^{-\bar{v}}x)$. First we
consider
\begin{eqnarray*}
  P(r' < x) &=& P(b < xe^{-v}) \\
  f_{r'}(x) &=& f_b(xe^{-v}) e^{-v}
\end{eqnarray*}
Averaging over all $v$, we get
\begin{equation}\label{eq:UncondPDFSymmetric}
  \begin{aligned}
    f_{r'}(x) =& \int_{-\infty}^{\infty} dv (2\pi\sigma^2)^{-1/2}
    e^{-v^2/2\sigma^2}(2\pi)^{-1/2} \exp(-x^2e^{-2v}/2) e^{-v} \\
    =& {1 \over 2\pi\sigma} \int_{-\infty}^{\infty} dv
    \exp\left(
      -{1 \over 2\sigma^2} v^2 - v -{1 \over 2} x^2 e^{-2v}
    \right)
  \end{aligned}
  \end{equation}
The last part of the integrand, $e^{-x^2 e^{-2v} / 2}$, is plotted in
figure \ref{fig:DoubleExp}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=85 252 100
  231]{../pics/DoubleExp.pdf}
  \caption{\small \it Plot of $\exp(-{1 \over 2} x^2 e^{-2v})$}
  \label{fig:DoubleExp}
\end{figure}
Therefore we make the following approximation:
\[
\exp\left(-{1 \over 2} x^2 e^{-2v}\right) \approx \left\{
  \begin{array}{lr}
    0 & \text{if } v < \ln|x| -{1 \over 2} \ln(2\ln 2) \\
    1 & \text{otherwise}
  \end{array}
\right.
\]
Here we note that $\exp\left(-{1 \over 2} x^2 e^{-2v}\right) = 1/2$ at
$v = \ln|x| -{1 \over 2} \ln(2\ln 2)$.

>>>>

The justification of modeling return series as independent,
identically distributed normal variates comes from imagining the
price process $S_t$ as a stochastic process with
  independent and normally distributed relative increments $dS_t / S_t
  = \sigma dw_t$, where $w_t$ represents the path of a geometric
  Brownian motion and $dw_t$ is normally distributed with mean 0 and
  variance $dt$.

By adding a drift term $\mu dt$ to allow a deterministic trend in the
relative price change, one can express the price $S_t$ as a stochastic
differential equation:
\begin{eqnarray*}
  {dS_t \over S_t} &=& \mu dt + \sigma dw_t \\
\end{eqnarray*}
By It\^o's lemma, a stochastic differential
equation for the logarithmic price $\ln S_t$ can be obtained
\begin{eqnarray*}
  d(\ln S_t) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
Solving this equation gives
\begin{eqnarray*}
  \ln S_t - \ln S_0 &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma w_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$ \cite{Bernt2000}. Therefore,
in a discrete-time model, where the length of each time step $\Delta
t$ is fixed, the return $r_t = \ln S(t) - \ln S(t - \Delta t)$ is
assumed to be Gaussian distributed and have mean and variance that are
functions of $\Delta t$.

@@

``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A \gls{garch}(p, q) model is defined by the following
equation system \cite{Bollerslev86}:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu_t + \epsilon_t \\
  \end{aligned}
\end{equation}
where $r_t$ is the return as mentioned earlier; $p$ and $q$ are
constant integers; $\mu_t$ is the mean process and considered a small
constant for intraday returns. $\epsilon_t$ is called the innovation
of the return and is a variate whose conditional distribution
\footnotemark is Gaussian and has variance $\sigma^2_t$:
\begin{eqnarray*}
  \sigma_t^2 &=& \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
  \sum_{i=1}^p \beta_i \sigma_{t-i}^2
\end{eqnarray*}
\footnotetext{
By conditional distribution we mean the distribution conditional on
the variate's history: The conditional distribution at time $t$ means
the distribution conditional on the history up to time $t-1$
in the discrete-time setting.
}
The quantity $\sigma_t$ introduced here is termed the volatility, and
its meaning and significance has been discussed in the introduction
\ref{chp:introduction}. In the model of geometric Brownian motion it
is assumed constant and hence simply denoted $\sigma$.

% The quantity $\sigma_t$ introduced here is of great importance in
% finance and termed the volatility. Statistically, it is the
% standard deviation of the distribution of $r_t$ and measures how
% volatile $r_t$ is. The volatility of an asset constitutes a major
% risk of investments on the asset and hence is very important.

@@

Figure \ref{fig:nordea_15min_y_qq} shows the normal probability plot of
$y_t$. It is evident from this figure that $y_t$ has fat tails.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=78 255 109
  123]{../pics/nordea2_y_normplot.pdf}
  \caption{\small \it Normal probability plot of residuals of
    log-volatility ($y_t$)}
  \label{fig:nordea_15min_y_qq}
\end{figure}
In addition, a simple calculation reveals that the distribution of
$y_t$ has skewness 0.2988 (shown in table
\ref{tab:nordea_15min_y_moments}).
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691 \\
    \hline
  \end{tabular}
  \caption{\small \it Moments of log-volatility residuals ($y_t$)}
  \label{tab:nordea_15min_y_moments}
\end{table}
Based on this information, we find that $y_t$ can be well
described by a Johnson Su distribution. This is a flexible
distribution that features fat tails and positive skewness. It is
defined by the following transformation \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  %\vspace{-18mm}
  \centering
    \includegraphics[scale=0.4, clip=true, trim=92 229 116
    133]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{\small \it Log-volatility residuals $y_t$ fitted to a
      Johnson Su distribution. Horizontal: values of $y_t$, denoted x;
      Vertical: $\ln\left(P(y_t < x)\right)$.}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}

@@

First of all, we look at the means and standard deviations of $\ln \sigma^F_t -
\ln \hat{\sigma}_t$, which are listed in table \ref{tab:nordea_2012}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    & \gls{sv} & \gls{garch} & Sample mean \\
    \hline
    $\E(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.0040 & -0.0008 &
    -0.2210 \\
    \hline
    $\text{std}(\ln \sigma^F_t - \ln \hat{\sigma}_t)$ & 0.2659 & 0.3011 &
    0.2893 \\
    \hline
  \end{tabular}
  \caption{\small \it Mean and standard Deviation of the forecasts'
    distribution}
  \label{tab:nordea_2012}
\end{table}
It is seen from table \ref{tab:nordea_2012} that, on average, the \gls{sv}
model over-estimates while \gls{garch} under-estimates. In terms of the
standard deviation of $\ln \sigma^F_t - \ln \hat{\sigma}_t$, the \gls{sv}
model wins with a small margin. In contrast, the sample mean forecast
clearly under-estimates the log-volatilities to a large
extent. So the efforts of building models have indeed led
  to more accurate forecasts.

@@

Another measure of the forecasts' quality can be the percentage of good
forecasts, where the criterion of ``good'' is defined, respectively,
as the forecast deviating less than 1\%, 5\%, or 10\% from the
corresponding realized log-volatility. Table
\ref{tab:nordea_2012_good} shows the respective percentage of the 3
kinds of forecasts. Again in this table it is seen that the \gls{sv}
model gives more accurate forecasts than does \gls{garch}. For
example, defining a ``good'' forecast as one that deviates less than
5\% from the realized volatility, the probability of obtaining such a
good forecast is 75\% using the \gls{sv} model, 71\% using the
\gls{garch} model, and only 53\% using the sample mean.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    ${|\ln \sigma^F_t - \ln \hat{\sigma}_t| \over |\ln \hat{\sigma}_t|}$
    & \gls{sv} & \gls{garch} & sample mean \\
    \hline
    1\% & 22\% & 14\% & 11\% \\
    \hline
    5\% & 75\% & 71\% & 53\% \\
    \hline
    10\% & 97\% & 94\% & 89\% \\
    \hline
  \end{tabular}
  % \begin{tabular}{|c|c|c|c|}
  %   \hline
  %   ${\ln \sigma^F_t - \ln \hat{\sigma}_t \over |\ln \hat{\sigma}_t|}$
  %   & \gls{sv} & GARCH & sample mean \\
  %   \hline
  %   1\% & 21.73\% & 13.83\% & 11.36\% \\
  %   \hline
  %   5\% & 74.57\% & 70.62\% & 53.33\% \\
  %   \hline
  %   10\% & 97.28\% & 93.83\% & 89.14\% \\
  %   \hline
  % \end{tabular}
  \caption{\small \it The percentage of ``good'' forecasts when the
    criterion of being good is deviating no more than 1\%, 5\% or
    10\%.}
  \label{tab:nordea_2012_good}
\end{table}

@@

\subsection{The Simplified model}\label{sec:SLV_Symmetric}
In the following we derive the unconditional \gls{pdf} of $r'$, the
de-meaned\footnote{By ``de-mean'' we mean subtracting from
  a random variable its expectation value.} and rescaled
return. Denote this \gls{pdf} $f_{r'}(x)$. Then the \gls{pdf} of $r$ is
$e^{-\bar{v}}f_{r'}(e^{-\bar{v}}x)$. First we consider
\begin{eqnarray*}
  P(r' < x) &=& P(b < xe^{-v}) \\
  f_{r'}(x) &=& f_b(xe^{-v}) e^{-v}
\end{eqnarray*}
Averaging over all $v$, we get
\begin{equation}\label{eq:UncondPDFSymmetric}
  \begin{aligned}
    f_{r'}(x) =& \int_{-\infty}^{\infty} dv (2\pi\sigma^2)^{-1/2}
    e^{-v^2/2\sigma^2}(2\pi)^{-1/2} \exp(-x^2e^{-2v}/2) e^{-v} \\
    =& {1 \over 2\pi\sigma} \int_{-\infty}^{\infty} dv
    \exp\left(
      -{1 \over 2\sigma^2} v^2 - v -{1 \over 2} x^2 e^{-2v}
    \right)
  \end{aligned}
  \end{equation}
The last part of the integrand, $e^{-x^2 e^{-2v} / 2}$, is plotted in
figure \ref{fig:DoubleExp}. We see from the figure that the landscape
defined by this function, roughly speaking, consists of two levels,
the lower one lying at height 0 and the higher one at height 1. The
slope connecting the two levels is rather steep. It changes as a
double exponential along lines of constant $x$ and as $e^{-x^2}$ along
lines of constant $v$. Therefore we approximate this function as a
2-dimensional step function:
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=85 252 100
  231]{../pics/DoubleExp.pdf}
  \caption{\small \it Plot of $\exp(-{1 \over 2} x^2 e^{-2v})$}
  \label{fig:DoubleExp}
\end{figure}
\[
\exp\left(-{1 \over 2} x^2 e^{-2v}\right) \approx \left\{
  \begin{array}{lr}
    0 & \text{if } v < \ln|x| -{1 \over 2} \ln(2\ln 2) \\
    1 & \text{otherwise}
  \end{array}
\right.
\]
Here we note that $\exp\left(-{1 \over 2} x^2 e^{-2v}\right) = 1/2$ at
$v = \ln|x| -{1 \over 2} \ln(2\ln 2)$.

### 2014-08-13 thesis/thesis.tex
Finalized thesis.

\documentclass{report}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{amssymb}
% \usepackage{amsfonts, amsmath}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage[acronym]{glossaries}
\usepackage{bookmark}

@@

% \title{Stocks' Cross-Correlations}
% \subtitle{}
\makeglossaries
\author{Xie Xiaolei}
\date{\today}
\begin{document}

@@

\end{center}
\end{titlepage}
% \maketitle
\begin{abstract}
Return models and covariance matrices of return series have been
studied. In particular, \gls{garch} and \gls{sv} models
are compared with respect to their forecasting accuracy when applied
to intraday return series. \gls{sv} models are found to be considerably more
accurate and more consistent in accuracy in forecasting.

Covariance matrices formed from Gaussian and \gls{garch} return series, and
in particular, return series auto-correlated as an AR(1) process, have
been studied. In the case of Gaussian returns, the largest eigenvalue
is found to approximately follow a gamma distribution also when the
returns are auto-correlated. Expressions relating the mean and the
variance of the asymptotic Gaussian distribution of the matrix
elements are derived. In the case of \gls{garch} returns, both the largest
and the smallest eigenvalues of the covariance matrix are seen to
increase with increasing auto-correlation. The matrix elements are
found to follow L\'evy distributions with different L\'evy indices for
the diagonal and the non-diagonal elements.

Localization of eigenvectors of covariance matrices of
returns from \gls{garch} processes has been investigated. It is
found that the localization is reduced as the auto-correlation is
increased. Quantitatively, the number of localized eigenvectors
decreases approximately as a quadratic function with the
auto-correlation strength, i.e. the autoregressive coefficient of the
AR(1) process.
\end{abstract}

\section*{Acknowledgements}
First of all, I thank my supervisor, Prof. Sven \AA berg, for the many
hours of mentoring, discussion, encouragement and guidance. Without
them, this report will not exist. Then I wish to give my gratitude to
the people in the division of mathematical physics. The casual
discussions that I had with them truely inspired me, and the friendly
and lively atmosphere that they create has helped bring peace and clarity
to my mind.

@@

\chapter{Introduction}
The mathematics of the financial market has always been a topic that
arouses interest and imagination, and with no doubt, has been
studied from many aspects and in many different ways.

@@

\Gls{sv} models built on historical data describe the time evolution
of the aforementioned observables in terms of probability
distributions, autocorrelations and cross correlations. They may have
some predictive power and help determine a fair price of a given
asset.

Quite often, instead of an asset's price, the relative price change,
i.e. the return, is studied. Analytically solvable models often assume
Gaussian return distribution, although data suggest fatter tails. In
numerical return models, realistic reproduction of historical data may
be achieved.

@@

This report is organized as follows: Chapter \ref{chp:PriceModels}
reviews some of the most influential return model. Parameters of the
models are fitted to a few intraday return series and the predictive
power of the models is compared. A calculation of the
unconditional distribution functions of \gls{sv} models,
especially in the case where the residual of the log-volatility
and the innovation of the return are correlated normal variates, is
also presented. In chapter \ref{chp:Gaussian} we investigate
distributions of eigenvalues of the Wishart matrix, and study the
influence of auto-correlated returns. In chapter
\ref{chp:CrossCorrelationFat} distributions of elements and
eigenvalues of the covariance matrix of identically specified
\gls{garch}(1,1) series are studied. Finally, chapter \ref{chp:summary}
summarizes the results. Supplementary materials are provided in the
appendices.

@@

\chapter{Self-reflection}
First of all, during the thesis project I learned a lot about models
of time series in finance, such as \gls{garch}, and \gls{sv} models,
as well as traditional time series analysis, for example, \gls{arima}
models. In addition, a study of continuous-time models gave me an
opportunity to increase my knowledge of stochastic differential
equations, probability theory, and statistics. I am a beginner, but
nonetheless find it comfortable to read books and articles on these
subjects after the thesis project.

Moreover, I have introduced myself to the random matrix theory,
multivariate analysis, and extreme value theory. These became relevant
when I worked on the covariance matrix of returns described by
Gaussian or \gls{garch} models. I am still a long way from being able
to do serious proofs or derivations in these areas, but I know what to
read and can understand the basics.

>>>>

\documentclass{report}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage[acronym]{glossaries}
\usepackage{bookmark}

@@

\makeglossaries
\author{Xie Xiaolei}
\date{\today}
\begin{document}

@@

\end{center}
\end{titlepage}
% \maketitle
\begin{abstract}
Return models and covariance matrices of return series have been
studied. In particular, \gls{garch} and \gls{sv} models
are compared with respect to their forecasting accuracy when applied
to intraday return series. \gls{sv} models are found to be considerably more
accurate and more consistent in accuracy of forecasting.

Covariance matrices formed from Gaussian and \gls{garch} return series, and
in particular, return series auto-correlated as an AR(1) process, have
been studied. In the case of Gaussian returns, the largest eigenvalue
is found to approximately follow a gamma distribution also when the
returns are auto-correlated. Expressions relating auto-correlation
strength with the mean and the variance of the asymptotic Gaussian
distribution of the matrix elements are derived. In the case of
\gls{garch} returns, both the largest and the smallest eigenvalues of
the covariance matrix are seen to increase with increasing
auto-correlation. The matrix elements are found to follow L\'evy
distributions with different L\'evy indices for the diagonal and the
non-diagonal elements.

Localization of eigenvectors of covariance matrices of
returns from \gls{garch} processes has been investigated. It is
found that the localization is reduced as the auto-correlation is
increased. Quantitatively, the number of localized eigenvectors
decreases approximately as a quadratic function with the
auto-correlation strength, i.e. the autoregressive coefficient of the
AR(1) process.

%% Popular Abstract
% Can we predict the movements of the financial market? The direct answer is "no", but there is more to be said. One cannot predict the price movement in an efficient market, but one can indeed predict how "volatile" the prices are (volatility). This does not lead to profits, but is useful for risk management --- if one pursues a profit, one must take a risk. Knowing and quantitatively managing the risks associated with investments, one often makes wiser decisions.

% This report entails, first of all, a comparison of return models in the context of intraday returns. 6 return series are studied: Nordea, Volvo and Ericsson in the time scales of 15 and 30 minutes. Stochastic Volatility (SV) models and GARCH models are fitted to these series and the resulting volatility forecasts are compared. SV models are found to be more accurate in most cases.

% Covariance matrices, which give the correlation between different companies, are also studied. In particular, the eigenvectors and eigenvalues of a covariance matrix, which respectively represent a set of driving factors of the companies' shares and the variances of these factors, are studied. It is found that when the return series are auto-correlated, i.e. when observations at different times are correlated, the eigenvectors are delocalized --- their expansion as a linear combination of the basis vectors involve more appreciable coefficients.
\end{abstract}

\section*{Acknowledgements}
First of all, I thank my supervisor, Prof. Sven \AA berg, for the many
hours of mentoring, discussion, encouragement and guidance. Without
them, this report will not exist. Then I wish to give my gratitude to
the people in the division of mathematical physics. The casual
discussions that I had with them truly inspired me, and
the friendly and lively atmosphere that they create has helped bring
peace and clarity to my mind.

@@

\chapter{Introduction} \label{chp:introduction}
The mathematics of the financial market has always been a topic that
arouses interest and imagination, and with no doubt, has been
studied from many aspects and in many different ways.

@@

Quite often, instead of an asset's price, the relative price change,
i.e. the return, is studied. Analytically solvable models often assume
Gaussian return distribution, although data suggest fatter tails
\footnotemark. By numerical methods, statistical features of
historical returns data, such as fat tails, can be reproduced.
\footnotetext{
  While the exact nature of the tails of the distribution of
  historical returns $r$ is debatable, it is agreed $P(|r| > x) > a
  e^{-bx^2}$, for any positive constants $a$ and $b$. Here $P(\cdot)$
  denotes probability. We adopt this as the definition of ``fat''
  tails.
}

Statistical models built on historical data describe the time
evolution of the aforementioned observables in terms of probability
distributions, autocorrelations and cross correlations. They have been
used to forecast future volatilities and help determine a pair price
of a contingent claim, e.g. an option. Here the term ``volatility''
refers to the standard deviation of the distribution of returns, and
measures how volatile the returns are. The volatility of an asset
constitutes a major risk of investments on the asset and hence is very
important.

@@

This report is organized as follows: Chapter \ref{chp:PriceModels}
reviews some of the most influential return model. Parameters of the
models are fitted to a few intraday return series and the predictive
power of the models is compared. A calculation of the
unconditional distribution functions of \gls{sv} models,
especially in the case where the residual of the log-volatility
and the innovation of the return are correlated normal variates, is
also presented. In chapter \ref{chp:Gaussian} we investigate
distributions of eigenvalues of the Wishart matrix, and study the
influence of auto-correlated returns. In chapter
\ref{chp:CrossCorrelationFat} distributions of elements and
eigenvalues of the covariance matrix of identically specified
\gls{garch}(1,1) series are studied. Finally, chapter \ref{chp:summary}
summarizes the results. Supplementary material is provided in the
appendices.

@@

\chapter{Self-reflection}
First of all, during the thesis project I learned a lot about models
of time series in finance, such as \gls{garch}, and \gls{sv} models,
as well as traditional time series analysis, for example, \gls{arima}
models. In addition, a study of continuous-time models gave me an
opportunity to increase my knowledge of stochastic processes,
probability theory and statistics.

Moreover, I have introduced myself to the random matrix theory,
multivariate analysis, and extreme value theory. These became relevant
when I worked on the covariance matrix of returns described by
Gaussian or \gls{garch} models.
